---
title: "ESM244_Lab1"
author: "Kristina Fauss"
date: "1/12/2022"
output: html_document
---

```{r setup, include=TRUE, warnings=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
#install.packages('tmap')
#install.packages('here')
#install.packages('sf')
#update.packages(ask=FALSE)
```

cmd-shift-enter : run cell

### Read in data

cmd-option-i : create code chunk
```{r}
sf_trees <- read_csv(here('data','sf_trees','sf_trees.csv'),
                     show_col_types=FALSE)
```

### Part 1: wrangling and ggplot

**EX 1** Find counts of obs by `legal_status` & wrangle

```{r}
### Method 1: group_by90 %>% summarize()
sf_trees %>%
  group_by(legal_status) %>%
  summarize(tree_count = n())

### Method 2: new way + functions
top_5_status <- sf_trees %>%
  count(legal_status) %>%
  drop_na(legal_status) %>% # drop any row with an na (in the legal status col)
  rename(tree_count=n) %>%
  relocate(tree_count) %>% # bring col to the front
  slice_max(tree_count, n=5) %>% # grab the top 5
  arrange(-tree_count) #low->high or high->low
```

Make a graph of top 5 from above

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y=tree_count))+ 
  geom_col(fill='dark green')+
  labs(x='Legal Status', y='Tree count')+
  coord_flip()+
  theme_dark()
```

**Ex 2** More wrangling: Only keep obs where legal status is "permitted site" and caretaker is `MTA`, and store as `permitted_data_df`

shift-cmd-c : comment out a line

```{r}
# sf_trees$legal_status %>% unique
# unique(sf_trees,$)
permitted_data_df <- sf_trees %>%
  filter(legal_status %in% c('Permitted Site','Private'), caretaker=='MTA') # , means 'and' can also use &; | means 'or'
  
```


**EX 3** Only keep Blackwood Acacia trees and then only keep columns `legal_status`, `date`, `latitude`, and `longitude` and store at `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>%
  filter(str_detect(species,'Blackwood Acacia')) %>% # as long at that str is in the element, it is kept (even if there's other stuff)
  select(legal_status, date, lat=latitude, lon=longitude) # lat=latitued renames lattitude to lat

### Make a graph of locations
ggplot(data=blackwood_acacia_df, aes(x=lat, y=lon))+
  geom_point(color='darkgreen')# this is not how you plot spatial data!
```

**EX 4** use `tidyr::seperate`

```{r}
sf_trees_sep <- sf_trees %>%
  separate(species, into=c('spp_scientific','spp_common'), sep=' :: ')
```

**EX 5** use `tidyr::unite`

```{r}
ex_5 <- sf_trees %>%
  unite('id_status', tree_id, legal_status, sep = '_COOL_') # 'id_status' is the new col name; sep is a seperator string you set
```














